---
title: "IE360 Homework 2"
author: "Fatih Mecidiye"     
output: html_document
date: '2022-05-07'
---

## Statistical Forecasting and Time Series

### Libraries

```{r warning=FALSE, message=FALSE}
require(data.table)
library(ggplot2)
library(dplyr)
require(skimr)
library(fpp)
require(ggcorrplot)
```

### Data Visualization

```{r,include=TRUE}
data_path <- "C:/Users/Asus/Desktop/2022 - Spring/IE360/HW2/IE360_Spring22_HW2_data.csv"

data <- fread(data_path)
colnames(data) <- c("Quarters", "UGS", "RNUV","NLPG","PU","PG","NUGV","NDGV","GNPA","GNPC","GNPT")

fmt <- "%Y_Q%q"
data$Quarters <- as.yearqtr(data$Quarters, format = fmt)

head(data)
str(data)
```

By looking the ggplot of UGS vs Quarters, we can say that the time series is not stationary.

```{r,include=TRUE, warning=FALSE, error=FALSE}
ggplot(data,aes(x=Quarters,y=UGS)) + geom_line()

ggplot(data[Quarters<"2002 Q2"],aes(x=Quarters,y=UGS)) + geom_line()
```

The ggplot of yearly UGS sale clearly demonstrates the decreasing trend.

```{r,include=TRUE,warning=FALSE, error=FALSE}
# showing decreasing trend yearly
data$year = as.numeric(format(data$Quarters, "%Y"))
yearly_UGS <- data[,list(yearly_sale = mean(UGS)),by= list(year)]
ggplot(yearly_UGS,aes(x=year,y=yearly_sale)) + geom_line()
```

### Model 1

To use decreasing trend component of the time series, I add the trend column to data.

```{r,include=TRUE}
# trend component added
data[,trend := 1: .N ]
head(data)
```

```{r,include=TRUE,warning=FALSE, error=FALSE}
# linear regression model with only trend
Model1 <- lm(UGS ~ trend, data)
summary(Model1)

data[,trend_const := predict(Model1,data)]

ggplot(data,aes(x=Quarters)) + geom_line(aes(y=UGS, color="real")) + geom_line(aes(y=trend_const, color="trend_const"))
```

### Model 2

Autocorrelation function of the UGS give us the information about the seasonality of our target variable.

```{r,include=TRUE}
acf(data$UGS[1:28])
```

By using the trend and seasonality component together in the regression model, we can get quite good model.

```{r,include=TRUE,warning=FALSE, error=FALSE}
data$QuarterFactor = as.factor(format(data$Quarters, "%q"))
Model2 <- lm(UGS ~ trend + QuarterFactor, data)
summary(Model2)
data[,trend_const_QuarterlySeasonality := predict(Model2,data)]
ggplot(data,aes(x=Quarters)) + geom_line(aes(y=UGS, color="real")) + geom_line(aes(y=trend_const_QuarterlySeasonality, color="trend_const_QuarterlySeasonality"))
```

```{r,include=TRUE,warning=FALSE, error=FALSE}
# residuals
ggplot(data,aes(x=Quarters)) + geom_line(aes(y=UGS - trend_const_QuarterlySeasonality, color="residual")) 
checkresiduals(Model2$residuals)
```

## Model 3 - Adding regressors to the model

```{r,include=TRUE}
# correlation variables
res_table <- data.table()
res_table <- data[,3:11]
res_table[,residuals := c(Model2$residuals, rep(NA,4))]


correl_info=cor(res_table[1:28])

ggcorrplot(correl_info, 
           hc.order = TRUE, 
           type = "lower",
           lab = TRUE)

```

By using the residuals of model2 (model with trend and seasonality), I look at the correlation between residuals and the independent variables given. As a result, I realize that the RNUV has the highest correlation with residuals.

```{r,include=TRUE,warning=FALSE, error=FALSE}
ggplot(res_table,aes(x=RNUV, y=residuals)) + geom_smooth()
```

```{r,include=TRUE,warning=FALSE, error=FALSE}
# Model with RNUV, trend, QuarterFactor(seasonality component)
Model3 <- lm(UGS ~ trend + QuarterFactor + RNUV, data)
summary(Model3)
data[,pred_model3 := predict(Model3,data)]
ggplot(data,aes(x=Quarters)) + geom_line(aes(y=UGS, color="real")) + geom_line(aes(y=pred_model3, color="model3"))
checkresiduals(Model3$residuals)
data$pred_model3[29:32]
```

### Model 4

```{r,include=TRUE}
# further improvements on model3
res_table[,residuals := c(Model3$residuals, rep(NA,4))]
res_table[,RNUV := NULL]

correl_info2=cor(res_table[1:28])

ggcorrplot(correl_info2, 
           hc.order = TRUE, 
           type = "lower",
           lab = TRUE)
```

The RNUV column is deleted and the residuals are updated respect to model3. Then, by looking correlation matrix and ggplots of residuals vs PU and PG, I choose second independent variable to use in the model as PG.

```{r,include=TRUE,warning=FALSE, error=FALSE}
ggplot(res_table,aes(x=PU, y=residuals)) + geom_smooth()
ggplot(res_table,aes(x=PG, y=residuals)) + geom_smooth()
```

In the model summary, F-statistic, p-value and adjusted R-squared values prove that the model4 is strong enough. Also, the residuals are close to be random as seen in the results of checkresiduals() function.

```{r,include=TRUE,warning=FALSE, error=FALSE}
Model4 <- lm(UGS ~ trend + QuarterFactor + RNUV + PG, data)
summary(Model4)
checkresiduals(Model4$residuals)
```

### Prediction

Using the Model4, I forecast unleaded gasoline sales for 2007 (all quarters).

```{r,include=TRUE,warning=FALSE, error=FALSE}
data[,pred_model4 := predict(Model4,data)]
ggplot(data,aes(x=Quarters)) + geom_line(aes(y=UGS, color="real")) + geom_line(aes(y=pred_model4, color="model4"))
prediction <- data.table()
prediction$Quarter <- data$Quarters[29:32]
prediction$prediction <- data$pred_model4[29:32]
prediction
```
